# Wind model

Work in progress, don't pay attention.

## Naive Model

Analagous to the temp model.

$$y_{ih} = \beta_0 + \beta_{1h} + \beta_{2m} + \beta_3*avg\_sfcWind + \beta_{4h}*avg\_sfcWind + \epsilon_i $$
The data was generated in the notebook `prepare_data.ipnyb`

```{r load_packages, results='hide', message=FALSE}
library("tidymodels")
library("data.table")
library("report")
library("gridExtra")
library("extremogram")
library("doParallel")
library("vip")
library("butcher")
source("utils.R")
source("metrics.R")

set.seed(1234) #TODO: Re run all with this seed
```

### Add last avg and next avg

```{r}
reanalysis <- fread('reanalysis_sfcWind_training_data.csv') |> mutate(month = as.factor(getMonth(time)),
                                                                      hour = as.factor(getHour(time)))
temp <- reanalysis |> mutate(date = getDate(time)) |> 
    group_by(date) |> summarize(sfcWind_daily = mean(sfcWind_daily)) |> 
    mutate(prev_sfcWind_daily = lag(sfcWind_daily), nxt_sfcWind_daily = lead(sfcWind_daily)) |>
    select(-sfcWind_daily) |>
    na.omit()
reanalysis <- reanalysis |> mutate(date = getDate(time)) |> inner_join(temp) |> select(-date)
rm(temp)
```

## Add the other variables

```{r}
temp_1 <- fread('reanalysis_tas_training_data.csv') |> 
          mutate(date = getDate(time)) |>
          group_by(date) |>
          summarize(tas_max = max(tas), tas_min = min(tas))
temperature <- fread('reanalysis_tas_training_data.csv') |> 
          mutate(date = getDate(time)) |>
          inner_join(temp_1, by=join_by(date)) |> select(-c(date,tas))

precipitation <- fread('reanalysis_precipitation_training_data.csv') |> filter(time < as.Date('2015-01-01')) |> select(-pr)

surface_pressure <- fread('reanalysis_surface_pressure_training_data.csv') |> 
  filter(time < as.Date('2015-01-01')) |> 
  select(-sp) |>
  mutate(sp_daily = sp_daily/100)

reanalysis <- reanalysis |> inner_join(precipitation) |> inner_join(surface_pressure) |> inner_join(temperature)
rm(precipitation, temp_1, temperature, surface_pressure)
```

```{r}
reanalysis_train <- reanalysis |> filter(time < as.Date('2014-01-01'))
reanalysis_test <- reanalysis |> filter(time >= as.Date('2014-01-01'))

reanalysis_reduced_train <- reanalysis_train |> select(-c(pr_daily,sp_daily,tas_daily,tas_max,tas_min))
reanalysis_reduced_test <- reanalysis_test |> select(-c(pr_daily,sp_daily,tas_daily,tas_max,tas_min))
```

```{r define_model, message=FALSE}
nv_recipe <- recipe(sfcWind ~ ., data = reanalysis_train) |>
                update_role(time, new_role = "ID") |>
                step_interact(~ hour:sfcWind_daily)

nv_model <- linear_reg() %>% set_engine("lm")

nv_workflow <- 
  workflow() %>% 
  add_model(nv_model) %>% 
  add_recipe(nv_recipe)

nv_fit <- fit(nv_workflow, reanalysis_train)

nv_pred <- predict(nv_fit, reanalysis_test)
rm(nv_recipe, nv_model, nv_workflow)
```

```{r, message=FALSE}
nv_reduced_recipe <- recipe(sfcWind ~ ., data = reanalysis_reduced_train) |>
                update_role(time, new_role = "ID") |>
                step_interact(~ hour:sfcWind_daily)

nv_reduced_model <- linear_reg() |> set_engine("lm")

nv_reduced_wf <- 
  workflow() |>
  add_model(nv_reduced_model) |>
  add_recipe(nv_reduced_recipe)

nv_reduced_fit <- fit(nv_reduced_wf, reanalysis_reduced_train)

nv_reduced_pred <- predict(nv_reduced_fit, reanalysis_reduced_test)
rm(nv_reduced_recipe, nv_reduced_model, nv_reduced_wf)
```

### KNNReg

```{r}
knnreg <- copy(reanalysis) |> filter(time = NA) |> select(-c(sfcWind_daily,hour)) #Copy dataframe for structure

for(i in 0:23){
  reanalysis_train_hour <- reanalysis_train |> filter(hour == i) |> select(-hour)
  reanalysis_test_hour <- reanalysis_test |> filter(hour == i) |> select(-hour)


  knnreg_recipe <- recipe(sfcWind ~ ., data = reanalysis_train_hour) |>
    step_scale(all_numeric_predictors()) |>
    step_center(all_numeric_predictors())

  knnreg_model <- nearest_neighbor(weight_func = "rectangular",
                                neighbors = floor(sqrt(nrow(reanalysis_train_hour)))) |>
                                set_engine("kknn") |>
                                set_mode("regression")

  knnreg_wf <- workflow() |>
    add_recipe(knnreg_recipe) |>
    add_model(knnreg_model)
  
  knnreg_fit <- fit(knnreg_wf, reanalysis_train_hour)
  
  reanalysis_test_hour$sfcWind <- knnreg_fit |>
                        predict(reanalysis_test_hour)
  reanalysis_test_hour <- reanalysis_test_hour |> select(-sfcWind_daily)
  knnreg <- knnreg |> bind_rows(reanalysis_test_hour)
}

knnreg_pred <- knnreg |> rename("knnreg_sfcWind" = "sfcWind") |> arrange(time) |> select(c(time, knnreg_sfcWind))
rm(knnreg_recipe, knnreg_model, knnreg_wf, reanalysis_test_hour, reanalysis_train_hour, knnreg_fit, knnreg)
```

```{r}
knnreg <- copy(reanalysis_reduced_train) |> filter(time = NA) |> select(-c(sfcWind_daily,hour)) #Copy dataframe for structure

for(i in 0:23){
  reanalysis_train_hour <- reanalysis_reduced_train |> filter(hour == i) |> select(-hour)
  reanalysis_test_hour <- reanalysis_reduced_test |> filter(hour == i) |> select(-hour)


  knnreg_reduced_recipe <- recipe(sfcWind ~ ., data = reanalysis_train_hour)  |>
    step_scale(all_numeric_predictors()) |>
    step_center(all_numeric_predictors())
  
  knnreg_reduced_model <- nearest_neighbor(weight_func = "rectangular",
                                neighbors = floor(sqrt(nrow(reanalysis_train_hour)))) |>
                                set_engine("kknn") |>
                                set_mode("regression")

  knnreg_reduced_wf <- workflow() |>
    add_recipe(knnreg_reduced_recipe) |>
    add_model(knnreg_reduced_model)
  
  knnreg_reduced_fit <- fit(knnreg_reduced_wf, reanalysis_train_hour)
  
  reanalysis_test_hour$sfcWind <- knnreg_reduced_fit |>
                        predict(reanalysis_test_hour)
  reanalysis_test_hour <- reanalysis_test_hour |> select(-sfcWind_daily)
  knnreg <- knnreg |> bind_rows(reanalysis_test_hour)
}

knnreg_reduced_pred <- knnreg |> rename("knnreg_reduced_sfcWind" = "sfcWind") |> arrange(time) |> select(c(time, knnreg_reduced_sfcWind))

rm(knnreg_reduced_recipe, knnreg_reduced_model, knnreg_reduced_wf, reanalysis_test_hour, reanalysis_train_hour, knnreg_reduced_fit, knnreg)
```

### XGBoost 
```{r}
xgb_recipe <- recipe(sfcWind ~ ., data = reanalysis_train) |>
                update_role(time, new_role = "ID") |>
                step_dummy(all_nominal()) 
                

xgb_model <- boost_tree(
  trees = 1000,
  tree_depth = tune(), min_n = tune(),
  loss_reduction = tune(),                     ## first three: model complexity
  sample_size = tune(), mtry = tune(),         ## randomness
  learn_rate = tune()                          ## step size
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_wf <- workflow() %>%
  add_recipe(xgb_recipe) %>%
  add_model(xgb_model)

xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), reanalysis_train),
  learn_rate(),
  size = 20
)

xgb_folds <- vfold_cv(reanalysis_train)

cl <- makeCluster(4)
registerDoParallel(cl)

xgb_tune_res <- tune_grid(
  xgb_wf,
  resamples = xgb_folds,
  grid = xgb_grid,
  control = control_grid(save_pred = TRUE)
)


### USE THE BEST XGBOOST
xgb_best <- select_best(xgb_tune_res, "rmse")

xgb_workflow <- xgb_wf |> finalize_workflow(xgb_best)

xgb_fit <- fit(xgb_workflow, reanalysis_train)

xgb_predicted <- predict(xgb_fit, reanalysis_test)

rm(xgb_wf, xgb_recipe, xgb_model, xgb_tune_res, xgb_folds, xgb_grid)
```

Save the best model of xgboost

```{r}
xgb_fit_butched <- butcher(xgb_fit)
saveRDS(xgb_fit_butched, file = "models/xgb_sfcWind.rds")
rm(xgb_fit_butched)
```


```{r}
xgb_reduced_recipe <- recipe(sfcWind ~ ., data = reanalysis_reduced_train) |>
                update_role(time, new_role = "ID") |>
                step_dummy(all_nominal()) 
                

xgb_reduced_model <- boost_tree(
  trees = 1000,
  tree_depth = tune(), min_n = tune(),
  loss_reduction = tune(),                     ## first three: model complexity
  sample_size = tune(), mtry = tune(),         ## randomness
  learn_rate = tune()                          ## step size
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_reduced_wf <- workflow() %>%
  add_recipe(xgb_reduced_recipe) %>%
  add_model(xgb_reduced_model)

xgb_reduced_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), reanalysis_reduced_train),
  learn_rate(),
  size = 20
)

xgb_reduced_folds <- vfold_cv(reanalysis_reduced_train)

cl <- makeCluster(4)
registerDoParallel(cl)

xgb_reduced_tune_res <- tune_grid(
  xgb_reduced_wf,
  resamples = xgb_reduced_folds,
  grid = xgb_reduced_grid,
  control = control_grid(save_pred = TRUE)
)


### USE THE BEST XGBOOST
xgb_reduced_best <- select_best(xgb_reduced_tune_res, "rmse")

xgb_reduced_wf <- xgb_reduced_wf |> finalize_workflow(xgb_reduced_best)

xgb_reduced_fit <- fit(xgb_reduced_wf, reanalysis_reduced_train)

xgb_reduced_predicted <- predict(xgb_reduced_fit, reanalysis_reduced_test)

rm(xgb_reduced_wf, xgb_reduced_recipe, xgb_reduced_model, xgb_reduced_tune_res, xgb_reduced_folds, xgb_reduced_grid)
```

## Results

Hours are statistically significant, as is their interaction with the sfcWind_dailyerature; months are not. 

```{r show_summary, echo=FALSE}
engine <- extract_fit_engine(nv_fit)
#Hours are statistically significant, as is their interaction with the sfcWind_dailyerature; months are not. 
engine |> report_table()
```

```{r show_summary_v2, echo=FALSE}
engine <- extract_fit_engine(nv_reduced_fit)
#Hours are statistically significant, as is their interaction with the sfcWind_dailyerature; months are not. 
engine |> report_table()
```

```{r}
xgb_engine <- extract_fit_engine(xgb_fit)
vip(xgb_engine)
```

```{r, echo=FALSE}
res <-  reanalysis_test |> rename("obs_sfcWind" = "sfcWind") |>
  cbind(nv_pred) |> rename("nv_sfcWind" = ".pred") |>
  cbind(nv_reduced_pred) |> rename("nv_reduced_sfcWind" = ".pred") |>
  inner_join(knnreg_pred, by = join_by(time)) |>
  inner_join(knnreg_reduced_pred, by = join_by(time)) |>
  cbind(xgb_predicted) |> rename("xgb_sfcWind" = ".pred") |>
  cbind(xgb_reduced_predicted) |> rename("xgb_reduced_sfcWind" = ".pred") |>
  select(c(time, obs_sfcWind, nv_sfcWind, nv_reduced_sfcWind, knnreg_sfcWind, knnreg_reduced_sfcWind, xgb_sfcWind, xgb_reduced_sfcWind))
```

```{r}
## Load the predicted value of the LSTM model
cnn_predicted <- fread('cnn_output_sfcWind.csv') |> 
                    rename("cnn_sfcWind" = "V1") #It's the one with the biggest lag-window

lstm_predicted <- fread('lstm_output_sfcWind.csv') |> 
                    slice(6:n()) |>
                    rename("lstm_sfcWind" = "V1")

res <- res |> 
    slice(11:nrow(res)) |> 
    cbind(lstm_predicted) |> 
    cbind(cnn_predicted) |>
    filter(time >= as.Date('2014-01-02'))
```

```{r}
## Load the predicted value of the LSTM model

res <- res |> cbind(cnn_predicted) |> rename("cnn_sfcWind" = "V1")  |> 
  filter(time >= as.Date('2014-01-02'))
```


Plot (hourly) of three days.

```{r show_plot, echo=FALSE}
res_to_plot <- res[0:72,] 

p1 <- ggplot(res_to_plot, aes(x=time)) +
  geom_line(aes(y=obs_sfcWind)) +
  geom_line(aes(y=nv_sfcWind), color = "red") + 
  labs(y = "sfcWind", x = "", title = "naive model")
  
p2 <- ggplot(res_to_plot, aes(x=time)) +
  geom_line(aes(y=obs_sfcWind)) +
  geom_line(aes(y=knnreg_sfcWind), color = "red") + 
  labs(y = "sfcWind", x = "", title = "knnreg model")

p3 <- ggplot(res_to_plot, aes(x=time)) +
  geom_line(aes(y=obs_sfcWind)) +
  geom_line(aes(y=xgb_sfcWind), color = "red") + 
  labs(y = "sfcWind", x = "", title = "xgboost model")

p4 <- ggplot(res_to_plot, aes(x=time)) +
  geom_line(aes(y=obs_sfcWind)) +
  geom_line(aes(y=lstm_sfcWind), color = "red") + 
  labs(y = "sfcWind", x = "", title = "lstm model")

p5 <- ggplot(res_to_plot, aes(x=time)) +
  geom_line(aes(y=obs_sfcWind)) +
  geom_line(aes(y=cnn_sfcWind), color = "red") + 
  labs(y = "sfcWind", x = "", title = "cnn model")

grid.arrange(p1, p2, p3, p4, p5, nrow = 3)
```

```{r, warning=FALSE}
r <- rbind(
        metrics(res$time, res$obs_sfcWind, res$nv_sfcWind, "nv"),
        metrics(res$time, res$obs_sfcWind, res$nv_reduced_sfcWind, "nv_reduced"),
        metrics(res$time, res$obs_sfcWind, res$knnreg_sfcWind, "knnreg"),
        metrics(res$time, res$obs_sfcWind, res$knnreg_reduced_sfcWind, "knnreg_reduced"),
        metrics(res$time, res$obs_sfcWind, res$xgb_sfcWind, "xgboost"),
        metrics(res$time, res$obs_sfcWind, res$xgb_reduced_sfcWind, "xgboost_reduced"),
        metrics(res$time, res$obs_sfcWind, res$lstm_sfcWind, "lstm"),
        metrics(res$time, res$obs_sfcWind, res$cnn_sfcWind, "cnn")
    )
r
```

```{r, warning=FALSE}
r <- rbind(
        metrics_2(res$time, res$obs_sfcWind, res$nv_sfcWind, "nv"),
        metrics_2(res$time, res$obs_sfcWind, res$nv_reduced_sfcWind, "nv_reduced"),
        metrics_2(res$time, res$obs_sfcWind, res$knnreg_sfcWind, "knnreg"),
        metrics_2(res$time, res$obs_sfcWind, res$knnreg_reduced_sfcWind, "knnreg_reduced"),
        metrics_2(res$time, res$obs_sfcWind, res$xgb_sfcWind, "xgboost"),
        metrics_2(res$time, res$obs_sfcWind, res$xgb_reduced_sfcWind, "xgboost_reduced"),
        metrics_2(res$time, res$obs_sfcWind, res$lstm_sfcWind, "lstm"),
        metrics_2(res$time, res$obs_sfcWind, res$cnn_sfcWind, "cnn")
    )
r
```

```{r}
p1 <- maximum_histograms(res$time, res$obs_sfcWind, res$nv_sfcWind) + scale_fill_discrete(labels = c("nv", "obs"))
p2 <- maximum_histograms(res$time, res$obs_sfcWind, res$knnreg_sfcWind) + scale_fill_discrete(labels = c("knnreg", "obs"))
p3 <- maximum_histograms(res$time, res$obs_sfcWind, res$xgb_sfcWind) + scale_fill_discrete(labels = c("xgboost", "obs"))
p4 <- maximum_histograms(res$time, res$obs_sfcWind, res$xgb_reduced_sfcWind) + scale_fill_discrete(labels = c("xg_reduced", "obs"))
p5 <- maximum_histograms(res$time, res$obs_sfcWind, res$lstm) + scale_fill_discrete(labels = c("lstm", "obs"))
p6 <- maximum_histograms(res$time, res$obs_sfcWind, res$cnn) + scale_fill_discrete(labels = c("cnn", "obs"))

grid.arrange(p1, p2, p3, p4,p5, p6, nrow = 6, top = "How Often Peaks Hit Hourly")
```

```{r}
res_to_plot <- res |> 
    select(!ends_with("sfcWind_daily")) |>
    pivot_longer(
     cols = ends_with("_sfcWind"),
     names_to = "model",
     names_pattern = "(.*)_sfcWind",
     values_to = "sfcWind",
     values_drop_na = TRUE
    ) 
#|> filter(model %in% c('obs', 'lstm', 'xgb'))
  
ggplot(res_to_plot, aes(x=sfcWind, color = model))+
  geom_density()
```
```{r}
# Function to create a Q-Q plot comparing quantiles of a sample column against a reference column
qq_plot_against <- function(data, sample_col) {
  ref_col <- "obs_sfcWind"
  sample_data <- data[[sample_col]]
  ref_data <- data[[ref_col]]
  
  # Sort the data
  sample_data_sorted <- sort(sample_data)
  ref_data_sorted <- sort(ref_data)
  
  # Calculate quantiles
  n <- min(length(sample_data_sorted), length(ref_data_sorted))
  quantiles_sample <- sample_data_sorted[seq(1, length(sample_data_sorted), length.out = n)]
  quantiles_ref <- ref_data_sorted[seq(1, length(ref_data_sorted), length.out = n)]
  
  # Create a data frame for plotting
  quantile_df <- data.frame(Quantiles_Ref = quantiles_ref, Quantiles_Sample = quantiles_sample)
  
  # Generate the Q-Q plot
  ggplot(quantile_df, aes(x = Quantiles_Ref, y = Quantiles_Sample)) +
    geom_point() +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    ggtitle(paste("Q-Q Plot of", sample_col, "against", ref_col)) +
    xlab(paste("Quantiles of", ref_col)) +
    ylab(paste("Quantiles of", sample_col)) +
    theme_minimal()
}

# Columns to compare against the reference column
columns <- c("nv_sfcWind", "xgb_sfcWind", "lstm_sfcWind", "cnn_sfcWind")

# Generate Q-Q plots
plots <- lapply(columns, function(col) qq_plot_against(res, col))

# Arrange the plots using gridExtra
do.call(grid.arrange, c(plots, ncol = 2))  # Arrange in a grid
```
## Extremograms

```{r}
p1 <- extremogram1(res$obs_sfcWind, quant = .97, maxlag = 48, type = 1, ploting = 0)
p2 <- extremogram1(res$nv_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p3 <- extremogram1(res$knnreg_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p4 <- extremogram1(res$xgb_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p5 <- extremogram1(res$nv_reduced_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p6 <- extremogram1(res$knnreg_reduced_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p7 <- extremogram1(res$xgb_reduced_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p8 <- extremogram1(res$lstm_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p9 <- extremogram1(res$cnn_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)

res_to_plot <- data.frame(model = c(rep("obs", 48), rep("nv", 48), rep("knnreg", 48), rep("xgboost", 48),
                                    rep("nv_reduced", 48), rep("knnreg_reduced", 48), rep("xgboost_reduced", 48),
                                    rep("lstm",48), rep("cnn",48)),
                          lag = c(1:48, 1:48, 1:48, 1:48, 1:48, 1:48, 1:48, 1:48, 1:48),
                          extremogram = c(p1,p2,p3,p4,p5,p6,p7,p8,p9))

ggplot(res_to_plot, mapping = aes(x = lag, y = extremogram)) +
    geom_segment(mapping = aes(xend = lag, yend = 0)) +
    facet_wrap(~ model, nrow = 2)
```

```{r}
p1 <- acf(res$obs_sfcWind, lag.max = 47, plot = F)
p2 <- acf(res$nv_sfcWind, lag.max = 47, plot = F)
p3 <- acf(res$knnreg_sfcWind, lag.max = 47, plot = F)
p4 <- acf(res$xgb_sfcWind, lag.max = 47, plot = F)
p5 <- acf(res$nv_reduced_sfcWind, lag.max = 47, plot = F)
p6 <- acf(res$knnreg_reduced_sfcWind, lag.max = 47, plot = F)
p7 <- acf(res$xgb_reduced_sfcWind, lag.max = 47, plot = F)
p8 <- acf(res$lstm_sfcWind, lag.max = 47, plot = F)
p9 <- acf(res$cnn_sfcWind, lag.max = 47, plot = F)

res_to_plot <- data.frame(model = c(rep("obs", 48), rep("nv", 48), rep("knnreg", 48), rep("xgboost", 48),
                                    rep("nv_reduced", 48), rep("knnreg_reduced", 48), rep("xgboost_reduced", 48), 
                                    rep("lstm", 48), rep("cnn", 48)),
                          lag = c(1:48, 1:48, 1:48, 1:48, 1:48, 1:48, 1:48, 1:48, 1:48),
                          acf = c(p1$acf,p2$acf,p3$acf,p4$acf,p5$acf, p6$acf, p7$acf, p8$acf, p9$acf))

ggplot(res_to_plot, mapping = aes(x = lag, y = acf)) +
    geom_segment(mapping = aes(xend = lag, yend = 0)) +
    facet_wrap(~ model, nrow = 2)
```
