# Wind model

Work in progress, don't pay attention.

## Naive Model

Analagous to the temp model.

$$y_{ih} = \beta_0 + \beta_{1h} + \beta_{2m} + \beta_3*avg\_sfcWind + \beta_{4h}*avg\_sfcWind + \epsilon_i $$
The data was generated in the notebook `prepare_data.ipnyb`

```{r load_packages, results='hide', message=FALSE}
library("tidymodels")
library("data.table")
library("report")
library("gridExtra")
library("extremogram")
library("doParallel")
source("utils.R")
source("metrics.R")

reanalysis <- fread('reanalysis_sfcWind_training_data.csv') |> mutate(month = as.factor(getMonth(time)),
                                                                      hour = as.factor(getHour(time)))
```

### Add last avg and next avg

```{r}
temp <- reanalysis |> mutate(date = getDate(time)) |> 
    group_by(date) |> summarize(avg_sfcWind = mean(avg_sfcWind)) |> 
    mutate(prev_avg_sfcWind = lag(avg_sfcWind), nxt_avg_sfcWind = lead(avg_sfcWind)) |>
    select(-avg_sfcWind) |>
    na.omit()
reanalysis <- reanalysis |> mutate(date = getDate(time)) |> inner_join(temp) |> select(-date)
rm(temp)
```

## Add the other variables

```{r}
temp_1 <- fread('reanalysis_tas_training_data.csv') |> 
          mutate(date = getDate(time)) |>
          group_by(date) |>
          summarize(tas_max = max(tas), tas_min = min(tas))
temperature <- fread('reanalysis_tas_training_data.csv') |> 
          mutate(date = getDate(time)) |>
          inner_join(temp_1, by=join_by(date)) |> select(-date)

precipitation <- fread('reanalysis_precipitation_training_data.csv') |> filter(time < as.Date('2015-01-01')) |> select(-pr)

surface_pressure <- fread('reanalysis_surface_pressure_training_data.csv') |> 
  filter(time < as.Date('2015-01-01')) |> 
  select(-sp) |>
  mutate(sp_daily = sp_daily/100)

reanalysis <- reanalysis |> inner_join(precipitation) |> inner_join(surface_pressure) |> inner_join(temperature)
rm(precipitation, temp_1, temperature, surface_pressure)
```

```{r}
reanalysis_train <- reanalysis |> filter(time < as.Date('2014-01-01'))
reanalysis_test <- reanalysis |> filter(time >= as.Date('2014-01-01'))

reanalysis_reduced_train <- reanalysis_train |> select(-c(pr_daily,sp_daily,avg_tas,tas,tas_max,tas_min))
reanalysis_reduced_test <- reanalysis_test |> select(-c(pr_daily,sp_daily,avg_tas,tas,tas_max,tas_min))
```

```{r define_model, message=FALSE}
nv_recipe <- recipe(sfcWind ~ ., data = reanalysis_train) |>
                update_role(time, new_role = "ID") |>
                step_interact(~ hour:avg_sfcWind)

nv_model <- linear_reg() %>% set_engine("lm")

nv_workflow <- 
  workflow() %>% 
  add_model(nv_model) %>% 
  add_recipe(nv_recipe)

nv_fit <- fit(nv_workflow, reanalysis_train)

nv_pred <- predict(nv_fit, reanalysis_test)
rm(nv_recipe, nv_model, nv_workflow)
```

```{r, message=FALSE}
nv_reduced_recipe <- recipe(sfcWind ~ ., data = reanalysis_reduced_train) |>
                update_role(time, new_role = "ID") |>
                step_interact(~ hour:avg_sfcWind)

nv_reduced_model <- linear_reg() |> set_engine("lm")

nv_reduced_wf <- 
  workflow() |>
  add_model(nv_reduced_model) |>
  add_recipe(nv_reduced_recipe)

nv_reduced_fit <- fit(nv_reduced_wf, reanalysis_reduced_train)

nv_reduced_pred <- predict(nv_reduced_fit, reanalysis_reduced_test)
rm(nv_reduced_recipe, nv_reduced_model, nv_reduced_wf)
```

### KNNReg

```{r}
knnreg <- copy(reanalysis) |> filter(time = NA) |> select(-c(avg_sfcWind,hour)) #Copy dataframe for structure

for(i in 0:23){
  reanalysis_train_hour <- reanalysis_train |> filter(hour == i) |> select(-hour)
  reanalysis_test_hour <- reanalysis_test |> filter(hour == i) |> select(-hour)


  knnreg_recipe <- recipe(sfcWind ~ ., data = reanalysis_train_hour) |>
    step_scale(all_numeric_predictors()) |>
    step_center(all_numeric_predictors())

  knnreg_model <- nearest_neighbor(weight_func = "rectangular",
                                neighbors = floor(sqrt(nrow(reanalysis_train_hour)))) |>
                                set_engine("kknn") |>
                                set_mode("regression")

  knnreg_wf <- workflow() |>
    add_recipe(knnreg_recipe) |>
    add_model(knnreg_model)
  
  knnreg_fit <- fit(knnreg_wf, reanalysis_train_hour)
  
  reanalysis_test_hour$sfcWind <- knnreg_fit |>
                        predict(reanalysis_test_hour)
  reanalysis_test_hour <- reanalysis_test_hour |> select(-avg_sfcWind)
  knnreg <- knnreg |> bind_rows(reanalysis_test_hour)
}

knnreg_pred <- knnreg |> rename("knnreg_sfcWind" = "sfcWind") |> arrange(time) |> select(c(time, knnreg_sfcWind))
rm(knnreg_recipe, knnreg_model, knnreg_wf, reanalysis_test_hour, reanalysis_train_hour, knnreg_fit, knnreg)
```

```{r}
knnreg <- copy(reanalysis_reduced_train) |> filter(time = NA) |> select(-c(avg_sfcWind,hour)) #Copy dataframe for structure

for(i in 0:23){
  reanalysis_train_hour <- reanalysis_reduced_train |> filter(hour == i) |> select(-hour)
  reanalysis_test_hour <- reanalysis_reduced_test |> filter(hour == i) |> select(-hour)


  knnreg_reduced_recipe <- recipe(sfcWind ~ ., data = reanalysis_train_hour)  |>
    step_scale(all_numeric_predictors()) |>
    step_center(all_numeric_predictors())
  
  knnreg_reduced_model <- nearest_neighbor(weight_func = "rectangular",
                                neighbors = floor(sqrt(nrow(reanalysis_train_hour)))) |>
                                set_engine("kknn") |>
                                set_mode("regression")

  knnreg_reduced_wf <- workflow() |>
    add_recipe(knnreg_reduced_recipe) |>
    add_model(knnreg_reduced_model)
  
  knnreg_reduced_fit <- fit(knnreg_reduced_wf, reanalysis_train_hour)
  
  reanalysis_test_hour$sfcWind <- knnreg_reduced_fit |>
                        predict(reanalysis_test_hour)
  reanalysis_test_hour <- reanalysis_test_hour |> select(-avg_sfcWind)
  knnreg <- knnreg |> bind_rows(reanalysis_test_hour)
}

knnreg_reduced_pred <- knnreg |> rename("knnreg_reduced_sfcWind" = "sfcWind") |> arrange(time) |> select(c(time, knnreg_sfcWind))

rm(knnreg_reduced_recipe, knnreg_reduced_model, knnreg_reduced_wf, reanalysis_test_hour, reanalysis_train_hour, knnreg_reduced_fit, knnreg)
```

### XGBoost 
```{r}
xgb_recipe <- recipe(sfcWind ~ ., data = reanalysis_train) |>
                update_role(time, new_role = "ID") |>
                step_dummy(all_nominal()) 
                

xgb_model <- boost_tree(
  trees = 1000,
  tree_depth = tune(), min_n = tune(),
  loss_reduction = tune(),                     ## first three: model complexity
  sample_size = tune(), mtry = tune(),         ## randomness
  learn_rate = tune()                          ## step size
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_wf <- workflow() %>%
  add_recipe(xgb_recipe) %>%
  add_model(xgb_model)

xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), reanalysis_train),
  learn_rate(),
  size = 20
)

xgb_folds <- vfold_cv(reanalysis_train)

cl <- makeCluster(4)
registerDoParallel(cl)

xgb_tune_res <- tune_grid(
  xgb_wf,
  resamples = xgb_folds,
  grid = xgb_grid,
  control = control_grid(save_pred = TRUE)
)


### USE THE BEST XGBOOST
xgb_best <- select_best(xgb_tune_res, "rmse")

xgb_workflow <- xgb_wf |> finalize_workflow(xgb_best)

xgb_fit <- fit(xgb_workflow, reanalysis_train)

xgb_predicted <- predict(xgb_fit, reanalysis_test)

rm(xgb_wf, xgb_recipe, xgb_model, xgb_tune_res, xgb_folds, xgb_grid)
```

```{r}
xgb_reduced_recipe <- recipe(sfcWind ~ ., data = reanalysis_reduced_train) |>
                update_role(time, new_role = "ID") |>
                step_dummy(all_nominal()) 
                

xgb_reduced_model <- boost_tree(
  trees = 1000,
  tree_depth = tune(), min_n = tune(),
  loss_reduction = tune(),                     ## first three: model complexity
  sample_size = tune(), mtry = tune(),         ## randomness
  learn_rate = tune()                          ## step size
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_reduced_wf <- workflow() %>%
  add_recipe(xgb_reduced_recipe) %>%
  add_model(xgb_reduced_model)

xgb_reduced_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), reanalysis_reduced_train),
  learn_rate(),
  size = 20
)

xgb_reduced_folds <- vfold_cv(reanalysis_reduced_train)

cl <- makeCluster(4)
registerDoParallel(cl)

xgb_reduced_tune_res <- tune_grid(
  xgb_reduced_wf,
  resamples = xgb_reduced_folds,
  grid = xgb_reduced_grid,
  control = control_grid(save_pred = TRUE)
)


### USE THE BEST XGBOOST
xgb_reduced_best <- select_best(xgb_reduced_tune_res, "rmse")

xgb_reduced_wf <- xgb_reduced_wf |> finalize_workflow(xgb_reduced_best)

xgb_reduced_fit <- fit(xgb_reduced_wf, reanalysis_reduced_train)

xgb_reduced_predicted <- predict(xgb_reduced_fit, reanalysis_reduced_test)

rm(xgb_reduced_wf, xgb_reduced_recipe, xgb_reduced_model, xgb_reduced_tune_res, xgb_reduced_folds, xgb_reduced_grid)
```

## Results

Hours are statistically significant, as is their interaction with the avg_sfcWinderature; months are not. 

```{r show_summary, echo=FALSE}
engine <- extract_fit_engine(model_fit)
#Hours are statistically significant, as is their interaction with the avg_sfcWinderature; months are not. 
engine |> report_table()
```

```{r show_summary_v2, echo=FALSE}
engine <- extract_fit_engine(model_fit_v2)
#Hours are statistically significant, as is their interaction with the avg_sfcWinderature; months are not. 
engine |> report_table()
```

```{r, echo=FALSE}
res <-  reanalysis_test |> rename("obs_sfcWind" = "sfcWind") |>
  cbind(nv_pred) |> rename("nv_sfcWind" = ".pred") |>
  cbind(nv_reduced_pred) |> rename("nv_reduced_sfcWind" = ".pred") |>
  inner_join(knnreg_pred, by = join_by(time)) |>
  inner_join(knnreg_reduced_pred, by = join_by(time)) |>
  cbind(xgb_predicted) |> rename("xgb_sfcWind" = ".pred") |>
  cbind(xgb_reduced_predicted) |> rename("xgb_reduced_sfcWind" = ".pred") |>
  select(c(time, obs_sfcWind, nv_sfcWind, nv_reduced_sfcWind, knnreg_sfcWind, knnreg_reduced_sfcWind, xgb_sfcWind, xgb_reduced_sfcWind))
```

Plot (hourly) of three days.

```{r show_plot, echo=FALSE}
res_to_plot <- res[0:72,] 

p1 <- ggplot(res_to_plot, aes(x=time)) +
  geom_line(aes(y=obs_sfcWind)) +
  geom_line(aes(y=nv_sfcWind), color = "red") + 
  labs(y = "sfcWind", x = "", title = "naive model")
  
p2 <- ggplot(res_to_plot, aes(x=time)) +
  geom_line(aes(y=obs_sfcWind)) +
  geom_line(aes(y=knnreg_sfcWind), color = "red") + 
  labs(y = "sfcWind", x = "", title = "knnreg model")

p3 <- ggplot(res_to_plot, aes(x=time)) +
  geom_line(aes(y=obs_sfcWind)) +
  geom_line(aes(y=xgb_sfcWind), color = "red") + 
  labs(y = "sfcWind", x = "", title = "xgboost model ")

grid.arrange(p1, p2, p3, nrow = 2)
```

```{r, warning=FALSE}
r <- rbind(
        metrics(res$time, res$obs_sfcWind, res$nv_sfcWind, "nv"),
        metrics(res$time, res$obs_sfcWind, res$nv_reduced_sfcWind, "nv_reduced"),
        metrics(res$time, res$obs_sfcWind, res$knnreg_sfcWind, "knnreg"),
        metrics(res$time, res$obs_sfcWind, res$knnreg_reduced_sfcWind, "knnreg_reduced"),
        metrics(res$time, res$obs_sfcWind, res$xgb_sfcWind, "xgboost"),
        metrics(res$time, res$obs_sfcWind, res$xgb_reduced_sfcWind, "xgboost_reduced")
    )
r
```

```{r}
p1 <- maximum_histograms(res$time, res$obs_sfcWind, res$nv_sfcWind) + scale_fill_discrete(labels = c("nv", "obs"))
p2 <- maximum_histograms(res$time, res$obs_sfcWind, res$knnreg_sfcWind) + scale_fill_discrete(labels = c("knnreg", "obs"))
p3 <- maximum_histograms(res$time, res$obs_sfcWind, res$xgb_sfcWind) + scale_fill_discrete(labels = c("xgboost", "obs"))
p4 <- maximum_histograms(res$time, res$obs_sfcWind, res$xgb_reduced_sfcWind) + scale_fill_discrete(labels = c("xg_reduced", "obs"))

grid.arrange(p1, p2, p3, p4, nrow = 4, top = "How Often Peaks Hit Hourly")
```

## Extremograms

```{r}
res_to_plot <- res |> 
    select(!ends_with("avg_sfcWind")) |>
    pivot_longer(
     cols = ends_with("_sfcWind"),
     names_to = "model",
     names_pattern = "(.*)_sfcWind",
     values_to = "sfcWind",
     values_drop_na = TRUE
    ) 
  
ggplot(res_to_plot, aes(x=sfcWind, color = model))+
  geom_density()
```

```{r}
p1 <- extremogram1(res$obs_sfcWind, quant = .97, maxlag = 48, type = 1, ploting = 0)
p2 <- extremogram1(res$nv_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p3 <- extremogram1(res$knnreg_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p4 <- extremogram1(res$xgb_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p5 <- extremogram1(res$nv_reduced_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p6 <- extremogram1(res$knnreg_reduced_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)
p7 <- extremogram1(res$xgb_reduced_sfcWind, quant = .97, maxlag = 48, type = 1,  ploting = 0)

res_to_plot <- data.frame(model = c(rep("obs", 48), rep("nv", 48), rep("knnreg", 48), rep("xgboost", 48),
                                    rep("nv_reduced", 48), rep("knnreg_reduced", 48), rep("xgboost_reduced", 48)),
                          lag = c(1:48, 1:48, 1:48, 1:48, 1:48, 1:48, 1:48),
                          extremogram = c(p1,p2,p3,p4,p5,p6,p7))

ggplot(res_to_plot, mapping = aes(x = lag, y = extremogram)) +
    geom_segment(mapping = aes(xend = lag, yend = 0)) +
    facet_wrap(~ model, nrow = 2)
```

```{r}
p1 <- acf(res$obs_sfcWind, lag.max = 47, plot = F)
p2 <- acf(res$nv_sfcWind, lag.max = 47, plot = F)
p3 <- acf(res$knnreg_sfcWind, lag.max = 47, plot = F)
p4 <- acf(res$xgb_sfcWind, lag.max = 47, plot = F)
p5 <- acf(res$nv_reduced_sfcWind, lag.max = 47, plot = F)
p6 <- acf(res$knnreg_reduced_sfcWind, lag.max = 47, plot = F)
p7 <- acf(res$xgb_reduced_sfcWind, lag.max = 47, plot = F)

res_to_plot <- data.frame(model = c(rep("obs", 48), rep("nv", 48), rep("knnreg", 48), rep("xgboost", 48),
                                    rep("nv_reduced", 48), rep("knnreg_reduced", 48), rep("xgboost_reduced", 48)),
                          lag = c(1:48, 1:48, 1:48, 1:48, 1:48, 1:48, 1:48),
                          acf = c(p1$acf,p2$acf,p3$acf,p4$acf,p5$acf, p6$acf, p7$acf))

ggplot(res_to_plot, mapping = aes(x = lag, y = acf)) +
    geom_segment(mapping = aes(xend = lag, yend = 0)) +
    facet_wrap(~ model, nrow = 2)
```
