```{css}
h4 {
  margin-top: 3em;
  font-size: 18pt;
}
```

```{r}
colnames(res_validation_wide) <- sapply(strsplit(colnames(res_validation_wide), "\\."), `[`, 1)
colnames(res_validation_wide)[colnames(res_validation_wide) == "nv"] <- "naive"
```

#### Numerical evaluation metrics

- Each row correspond an evaluation metric
- Dark blue cell indicates the best model in the row. 
- Light blue cell indicates the second best model in the row.

```{r}
models <- res_validation_wide |> select(-c("time", "reanalysis", starts_with('undownscaled'))) |> colnames()

r <- lapply(models, function(x) {
    if (conf[["VARIABLES"]][[params$variable]][["daily"]]) {
      if(params$variable == 'pr'){
        metrics_unpaired_hourly_rain(time = res_validation_wide$time, 
                                truth = res_validation_wide$reanalysis, 
                                estimate = res_validation_wide[[x]], 
                                model = x)
      }else{
        metrics_unpaired_hourly(time = res_validation_wide$time, 
                                truth = res_validation_wide$reanalysis, 
                                estimate = res_validation_wide[[x]], 
                                model = x)        
      }
    } else {
      metrics_unpaired_daily(time = res_validation_wide$time, 
                             truth = res_validation_wide$reanalysis, 
                             estimate = res_validation_wide[[x]], model = x)
    }
})

results <- do.call(rbind, r) |> 
  arrange(abs(diff_of_means))
 # mutate(across(where(is.numeric), \(x) round(x, digits = 3) ) )

res.tabla <-  data.frame(Metric = names(results), t(results)) |> 
  mutate( sce = gsub('\\.', '_', res)) |> 
  remove_rownames()

# save numeric results 
# write.csv(res.tabla, 
#   paste0("reports/ds_acp/",params$variable,'_',gsub('\\.', '_', res), ".csv"), 
#   row.names = FALSE)  
#------------------------

# vector indicating if the metric is an error(-1), a gof (1), or a ratio (0)
metr.dir <- rep(1, length(names(results)) )
metr.dir[grep('mae|diff|ks', names(results))] <- -1
metr.dir[grep('ratio', names(results)) ] <- 0

data.frame(Metric = names(results), t(results)) |> 
  mutate( dr = metr.dir ) |>  # error direction
  as_tibble() |>  
  gt(  ) |>
  fmt_number(
    columns = everything(),
    decimals = 2
  ) |> 
  data_color(
    columns = -Metric, 
    method = 'numeric',
    direction = 'row',
    fn = fncol) |> 
      cols_hide(dr)



# res.tabla |>   as_tibble() |> 
#   select(-sce) |> 
#   gt() |> 
#   data_color(
#     columns = -c(Metric),
#     method = 'numeric',
#     direction = 'row',
#     palette = 'Blues', 
#     reverse=TRUE)
```

#### Time series of the first few days

```{r, echo=FALSE}
days <- n_distinct(res_validation_long$model)*n_distinct(res_validation_long$experiment)*24*5
res_to_plot <- res_validation_long[0:days,]


series_plot_ds(res_to_plot)
rm(res_validation_long)

######################################
# Figure of Temperature for final report 
# pp <- series_plot_ds(res_to_plot)
# built <- ggplot_build(pp)
# brk <- built$layout$panel_params[[1]]$x$breaks[2:4]
# pp + 
#   labs(x='', y = 'temperatura') + 
#   scale_y_continuous(
#     labels = function(x) round(x - 273.15),
#     name = "Temperatura"
#   ) +
#   scale_x_continuous(breaks = brk, 
#     labels = paste('Dia', c(3,5,7)) )
# ggsave(filename = 'reports/informes_aniiserie_ds_dias.png',  height = 6, width = 8)
######################################

```




#### Graphical evaluation metrics

<!-- *Important:* Right now we are only estimating the upper tail extremogram. Currently we didn't find a way to estimate the two tales at the same time. We are using `quant = .97` -->

- **Observations**: cumulative distribution (Detrended QQ), dependency structure (ACF), and extremes (Extremogram, using 0.97 cut off).

- **Daily/Monthly gruped**: Maximum hour histogram (only in daily variables), Amplitude distribution. 

- **Hourly/Daily grouped**: Mean profile per hour within days (or weeks witihin season)

- In all plots, grey color represent the true ERA5 signal. 
<br>

```{r}
plqq <- detrend_qq_ds(res_validation_wide)
```


```{r}
if (conf[["VARIABLES"]][[params$variable]][["daily"]]) {
  # res.mx <- res_validation_wide |> 
  #             select(-starts_with('undownscaled')) |>
  #             pivot_longer(-time, names_to = 'model', values_to = 'vs')  |> 
  #             mutate( date = ymd_hms(time) ) |> 
  #             mutate(dd = as_date(date), hh = hour(date)) |> 
  #             group_by(dd, model) |> 
  #             mutate(mx = max(vs) ) |> 
  #             ungroup() |> 
  #             filter(vs == mx) |> 
  #             select(model, dd, mx, hh)
  
  # pl.mx <- ggplot() +
  #   geom_bar(data=filter(res.mx, model =='reanalysis'), 
  #            mapping = aes(x=hh), fill='grey') +
  #   geom_freqpoly(data= filter(res.mx, model !='reanalysis'),
  #             mapping = aes(x=hh, color=model) ) +
  #   scale_color_brewer(palette = 'Dark2') + 
  #   theme_bw() 

  pl.mx <- res_validation_wide |> 
    select(-starts_with('undownscaled')) |> 
    hour_max_plot()
}
```

```{r}
#pl.hourly_distribution <- hourly_distribution(res_validation_wide |> select(-starts_with('undownscaled')))
if (conf[["VARIABLES"]][[params$variable]][["daily"]]) {
  pl.hourly_distribution <- res_validation_wide |> 
    select(-starts_with('undownscaled')) |> hourly_distribution()
} else { 
  pl.hourly_distribution <-   pl.hourly_distribution <- res_validation_wide |> 
    select(-starts_with('undownscaled')) |> week_distribution()
}
```

```{r}
if (conf[["VARIABLES"]][[params$variable]][["daily"]]) {
  pl.amplitude <- res_validation_wide |> 
    select(-starts_with('undownscaled')) |> amplitude_plot()
} else { 
  pl.amplitude <- res_validation_wide |> 
    select(-starts_with('undownscaled')) |> amplitude_monthly_plot()
}


#pl.amplitude <- amplitude_plot(res_validation_wide |> select(-starts_with('undownscaled')))
```

```{r}
pl.acf <- res_validation_wide |> select(-starts_with('undownscaled')) |> acf_plot()
```

```{r}
pl.ext <- res_validation_wide |> 
  select(-starts_with('undownscaled')) |> extremogram_plot()
```

```{r, fig.height=11, fig.width=11}
PP <- NULL
if (conf[["VARIABLES"]][[params$variable]][["daily"]]) {
  PP <- |> 
} else {
    PP <- (plqq | pl.ext | pl.acf) /
      (pl.amplitude) /
      (pl.hourly_distribution) + plot_layout(guides = 'collect')
}
PP
```