---
title: "`r paste0('Downscaling the variable ', params$variable)`"
params:
  variable: "clt" #Use pr as default
output:
  html_document:
    self_contained: true
---

```{r setup, include=FALSE}
library(here)
knitr::opts_knit$set(root.dir = here())
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

```{r imports_tab}
library(ggplot2)
library(tidyverse)
library(knitr)
library(gridExtra)
library(reticulate)
#use_python("/home/bruno/.miniconda3/envs/FSE_extremos/bin", required = TRUE)
library(yaml)
library(forcats)
library(patchwork)
library(gt)
library(plotly)
library("extremogram")
source('code/metrics.R')
source('code/utils.R')

#Load configuration
conf <- yaml.load_file("code/conf.yml")
```

```{r load_test_data_tab}
# params <- list(variables = 'clt')
x <- paste0('data/to_be_downscaled/', params$variable)
res_validation_wide_tot <- read.csv(paste0("data/downscaled_data/",params$variable,".csv"))
res_validation_long_tot <- res_validation_wide_tot |> 
                            pivot_longer(-c("time", "reanalysis"), names_to="model", values_to="value") |>
                            separate(col = model, sep = "\\.", into = c("model", "experiment"), extra = "merge")
to_be_downscaled <- res_validation_long_tot |> distinct(experiment) |> pull(experiment)
```

- Climate response variable: 
      - `sfcWind`, `tas`, `pr`: original scale *daily*, reduced scale: *hourly* 
      - `clt`, `rsds`: original scale *monthly*, reduced scale: *daily* 

- Training: 
      - Response: ERA5 in reduced scale 
      - Data perdiod: 1980 - 2015.  
      - Predictors: from CMIP6 (daily or monthly)
      - Predictors variables: `sfcWind`, `tas`, `pr`, `tasmax`, `tasmin` and `psl` in daily scale, `clt`, `rsdt`,`rsds` in monthly scale,  the month `r  if (conf[["VARIABLES"]][[params$variable]][["daily"]]) {", hour, sun's elevation & azimuth"}` and the daily daylight amount in seconds.

- Validation: 
    - Period:  `r min(res_validation_wide_tot$time)` to `r max(res_validation_wide_tot$time)`
    - Variable to predict: reduced scale CMIP6 variable  

<br> 

## Results {.tabset}

```{r tabs_tab, results='asis'}
# res <- to_be_downscaled[1]
for (res in to_be_downscaled) {
  
  # Subset the wide/long data for this experiment:
  res_validation_wide <- res_validation_wide_tot %>%
    select(time, reanalysis, ends_with(res))
  
  res_validation_long <- res_validation_long_tot %>%
    filter(experiment == res)
  
  # Emit the LEVEL‑3 heading for the new tab:
  cat("### ", res, "\n\n", sep = "")
  
  cat(
    knitr::knit_child(
      "code/downscaling_reports_generator/downscaling_reduced.Rmd",
      quiet = TRUE
    ),
    "\n\n",
    sep = ""
  )
}
```
